---
title: "Friedman test"
author: "Kristian Rubiano"
date: "30/08/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1 Testing for normality

### 1.1. Forest loss (%)  

Forest loss = ((forest_t0 - forest_t1)/forest_t0)*100

```{r}
library(car)
library(EnvStats)
library(nortest)
library(PMCMRplus)
library(rstatix)

data <- read.csv("maps_and_figures/boxplot_and_test/datos.csv")
knitr::kable(data)
```

```{r}
car::qqPlot(data$forest_loss_perc_p)
```

```{r}
plot(density(data$forest_loss_perc_p), lwd = 2, col = "red", main = "Density")
plot(density(data$forest_loss_perc_10), lwd = 2, col = "red", main = "Density")
plot(density(data$forest_loss_perc_20), lwd = 2, col = "red", main = "Density")
```

Shapiro-Wilks test: The test rejects the hypothesis of normality when the p-value is less than or equal to 0.05

```{r}
a <- shapiro.test(data$forest_loss_perc_p)
a$p.value < 0.05

b <- shapiro.test(data$forest_loss_perc_10)
b$p.value < 0.05

c <- shapiro.test(data$forest_loss_perc_20)
c$p.value < 0.05
```

Kolmogorov-Smirnov test: if p-values is less than 0.05 we reject the null hypotesis (the data do not come from a normal distribution)

```{r}
a <- ks.test(data$forest_loss_perc_p, "pnorm")
a$p.value < 0.05

b <- ks.test(data$forest_loss_perc_10, "pnorm")
b$p.value < 0.05

c <- ks.test(data$forest_loss_perc_20, "pnorm")
c$p.value < 0.05

```

Anderson-Darling test: The test rejects the hypothesis of normality when the p-value is less than or equal to 0.05.

```{r}
a <- nortest::ad.test(data$forest_loss_perc_p)
a$p.value < 0.05

b <- nortest::ad.test(data$forest_loss_perc_10)
b$p.value < 0.05

a <- nortest::ad.test(data$forest_loss_perc_20)
a$p.value < 0.05
```

Levene test: The test rejects the hypothesis of variance homogeneties when the p-value is less than or equal to 0.05.

```{r}
df_t <- as.data.frame(c(data$tnv_loss_perc_p, data$tnv_loss_perc_10, data$tnv_loss_perc_20))
names(df_t) <- "tnv_loss_perc"

df_t$forest_loss_perc <- c(data$forest_loss_perc_p, data$forest_loss_perc_10, data$forest_loss_perc_20)

df_t$onv_loss_perc <- c(data$onv_loss_perc_p, data$onv_loss_perc_10, data$onv_loss_perc_20)

df_t$group <- c(rep("p", length(data$tnv_loss_perc_p)), rep("b10", length(data$tnv_loss_perc_p)), rep("b20", length(data$tnv_loss_perc_p)))

df_t$group <- ordered(df_t$group, levels = c("p", "b10", "b20"))
levels(df_t$group)

df_t$park <- rep(1:51, 3)

a <- car::leveneTest(forest_loss_perc ~ group, df_t)
a$`Pr(>F)` < 0.05

```


### 1.2. Other natural vegetation loss (%)  

Other natural vegetation loss = ((onv_t0 - onv_t1)/onv_t0)*100

```{r}
car::qqPlot(data$onv_loss_perc_p)
car::qqPlot(data$onv_loss_perc_10)
car::qqPlot(data$onv_loss_perc_20)
```

```{r}
plot(density(data$onv_loss_perc_p), lwd = 2, col = "red", main = "Density")
plot(density(data$onv_loss_perc_10), lwd = 2, col = "red", main = "Density")
plot(density(data$onv_loss_perc_20), lwd = 2, col = "red", main = "Density")
```
Shapiro-Wilks test: The test rejects the hypothesis of normality when the p-value is less than or equal to 0.05


```{r}
a <- shapiro.test(data$onv_loss_perc_p)
a$p.value < 0.05

b <- shapiro.test(data$onv_loss_perc_10)
b$p.value < 0.05

c <- shapiro.test(data$onv_loss_perc_20)
c$p.value < 0.05
```

Kolmogorov-Smirnov test: if p-values is less than 0.05 we reject the null hypotesis (the data do not come from a normal distribution)

```{r}
a <- ks.test(data$onv_loss_perc_p, "pnorm")
a$p.value < 0.05

b <- ks.test(data$onv_loss_perc_10, "pnorm")
b$p.value < 0.05

c <- ks.test(data$onv_loss_perc_20, "pnorm")
c$p.value < 0.05

```

Anderson-Darling test: The test rejects the hypothesis of normality when the p-value is less than or equal to 0.05.

```{r}
a <- nortest::ad.test(data$onv_loss_perc_p)
a$p.value < 0.05

b <- nortest::ad.test(data$onv_loss_perc_10)
b$p.value < 0.05

c <- nortest::ad.test(data$onv_loss_perc_20)
c$p.value < 0.05
```

Levene test: The test rejects the hypothesis of variance homogeneties when the p-value is less than or equal to 0.05.

```{r}
a <- car::leveneTest(onv_loss_perc ~ group, df_t)
a$`Pr(>F)` < 0.05

```


### 1.3. Total natural vegetation loss (%)  

Total natural vegetation loss = ((tnv_t0 - tnv_t1)/tnv_t0)*100


```{r}
car::qqPlot(data$tnv_loss_perc_p)
car::qqPlot(data$tnv_loss_perc_10)
car::qqPlot(data$tnv_loss_perc_20)
```

```{r}
plot(density(data$tnv_loss_perc_p), lwd = 2, col = "red", main = "Density")
plot(density(data$tnv_loss_perc_10), lwd = 2, col = "red", main = "Density")
plot(density(data$tnv_loss_perc_20), lwd = 2, col = "red", main = "Density")
```

Shapiro-Wilks test: The test rejects the hypothesis of normality when the p-value is less than or equal to 0.05

```{r}
a <- shapiro.test(data$tnv_loss_perc_p)
a$p.value < 0.05

b <- shapiro.test(data$tnv_loss_perc_10)
b$p.value < 0.05

c <- shapiro.test(data$tnv_loss_perc_20)
c$p.value < 0.05
```

Kolmogorov-Smirnov test: if p-values is less than 0.05 we reject the null hypotesis (the data do not come from a normal distribution)

```{r}
a <- ks.test(data$tnv_loss_perc_p, "pnorm")
a$p.value < 0.05

b <- ks.test(data$tnv_loss_perc_10, "pnorm")
b$p.value < 0.05

c <- ks.test(data$tnv_loss_perc_20, "pnorm")
c$p.value < 0.05

```

Anderson-Darling test: The test rejects the hypothesis of normality when the p-value is less than or equal to 0.05.

```{r}
a <- nortest::ad.test(data$tnv_loss_perc_p)
a$p.value < 0.05

b <- nortest::ad.test(data$tnv_loss_perc_10)
b$p.value < 0.05

c <- nortest::ad.test(data$tnv_loss_perc_20)
c$p.value < 0.05
```

Levene test: The test rejects the hypothesis of variance homogeneties when the p-value is less than or equal to 0.05.

```{r}
a <- car::leveneTest(tnv_loss_perc ~ group, df_t)
a$`Pr(>F)` < 0.05

```


### 1.4. Conclusion  

Data seems not to follow a normal distribution. Then a non parametric test is needed.

Friedman test is chosen.

Armenteras, Rodriguez & Retana  (2009), Biological conservation, compared change rates whithin PAs and their buffers. They assumed the data to be paired.

Other test options can be found in Rodriguez, Armenteras & Retana (2013), Regional Environmental Change. They used repeated measures ANOVA, but it's a parametric test.

#### 2 Kruskal-Wallis and Friedman tests

### 2.1. Forest loss (%)

```{r}
a <- kruskal.test(forest_loss_perc ~ group, data = df_t)
a
a$p.value < 0.05

```

```{r}
pairwise.wilcox.test(df_t$forest_loss_perc, df_t$group, p.adjust.method = "BH")
```

```{r}
a <- friedman.test(forest_loss_perc ~ group | park, data = df_t)
a
a$p.value < 0.05
```


### 2.2. Other natural vegetation loss (%)

```{r}
a <- kruskal.test(onv_loss_perc ~ group, data = df_t)
a
a$p.value < 0.05

```

```{r}
pairwise.wilcox.test(df_t$onv_loss_perc, df_t$group, p.adjust.method = "BH")
```

```{r}
a <- friedman.test(onv_loss_perc ~ group | park, data = df_t)
a
a$p.value < 0.05
```


### 2.3. Total natural vegetation loss (%)

```{r}
a <- kruskal.test(tnv_loss_perc ~ group, data = df_t)
a
a$p.value < 0.05
```

```{r}
pairwise.wilcox.test(df_t$tnv_loss_perc, df_t$group, p.adjust.method = "BH")
```

```{r}
# data_tnv <- matrix(c(data$tnv_loss_perc_p, data$tnv_loss_perc_10, data$tnv_loss_perc_20), ncol = 3, nrow = 51, byrow = FALSE, dimnames = list(1:(51), c("p", "b1", "b2")))

PMCMRplus::friedmanTest(y = data_tnv, g = df_t$group, b = df_t$park)

ft <- friedman_test(formula = tnv_loss_perc ~ group | park, data = df_t)
# ft_label <- get_test_label(ft, detailed = TRUE)
ft



b <- create_test_label(n = 51,
  statistic.text = "χ2", statistic = 45.8,
  parameter = 2,
  p = "<0.0001",
  description = "Friedman test",
  type = "text",
  detailed = TRUE
)
b

get_pwc_label()

```

```{r}
fe <- friedman_effsize(formula = tnv_loss_perc ~ group | park, data = df_t, conf.level = 0.95, ci = TRUE)
fe

```

```{r}
# pairwise.wilcox.test(x = df_t$tnv_loss_perc, g = df_t$group, p.adjust.method = "bonferroni", paired = TRUE)
# PMCMRplus::frdAllPairsExactTest(y = df_t$tnv_loss_perc, g = df_t$group, b = df_t$park, p.adjust.method = "bonferroni")
# PMCMRplus::frdAllPairsNemenyiTest(y = df_t$tnv_loss_perc, g = df_t$group, b = df_t$park, p.adjust.method = "bonferroni")
# PMCMRplus::frdAllPairsSiegelTest(y = df_t$tnv_loss_perc, g = df_t$group, b = df_t$park, p.adjust.method = "bonferroni")
# PMCMRplus::frdAllPairsMillerTest(y = df_t$tnv_loss_perc, g = df_t$group, b = df_t$park, p.adjust.method = "bonferroni")

pht <- PMCMRplus::frdAllPairsConoverTest(y = df_t$tnv_loss_perc, g = df_t$group, b = df_t$park, p.adjust.method = "bonferroni")
pht

```

